---
title: 'Blog Post #1 - Introduction '
author: "Ellie Grueskin"
date: "2022-09-14"
output:
  html_document: default
  pdf_document: default
slug: []
categories: []
summary: In this post, I explore house wide races from 1948 to the present. I examine this primarily from a state-wide pooling with a special section at the end for the district I'll be following, Ohio-01.
tags: []
---

*Intro* 

Hello! My name is Ellie Grueskin, and I am a senior at Harvard College studying 
Economics. I will be logging my work in Government 1347: Election 
Analytics, taught by Ryan Enos, throughout the semester. In this introductory post, 
I analyze the popular vote shares in US House races over the last 70
years. At the end, I will also begin exploring my specific house race: Ohio-01.

*Data*

My dataset is a collection of the vote data for each house election, dating
back to 1948. Each observation contains the congressional district's name, state, candidates, and respective votes for each election. As well, it gives information as to whether
the candidate was an incumbent or challenger. In this analysis, I reformat the data
to exclude independent and write-in candidates, such that the 'vote share' margin is  
calculated as the number of Democratic or Republican votes divided by the sum of 
their votes. 

*Results - Swing States over Time* 

In my first plot, I examine the four-year change in share of a state's housewide popular votes that went to Republicans. I start with 1952 because the dataset only started in 1948. To clarify, a Republican vote margin above 50% means that Republicans received the majority 
of the popular vote in the state during that year. This does not necessarily translate 
to house seats, as districts are not divided evenly across state's party divisions.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
h<-read.csv("Week1_data/house party vote share by district 1948-2020.csv")
library(tidyverse)
library(usmap)
library(geomtextpath)

#Find house averages by state and year
h_avg <- h %>%
  select(raceYear, State,  RepVotes, DemVotes) %>%
# summarize party vote share by state
  group_by(State,raceYear) %>%
# mutate Rep vote margin by state %>%
  summarise(R_votemargin_st = (sum(RepVotes))/
           sum(RepVotes + DemVotes),
D_votemargin_st = (sum(DemVotes))/
sum(RepVotes + DemVotes)) %>%
rename(state = State)

h_avg_2<-h_avg %>%
  group_by(state) %>%
  #Lag by 4 years or 2 observations
  mutate(diff_4yr = R_votemargin_st-lag(R_votemargin_st,2),
         diff_2yr= R_votemargin_st-lag(R_votemargin_st))
```

```{r, message=FALSE}



# 4- year differences
p_4yr<-plot_usmap(data=h_avg_2%>% filter(raceYear>1950), regions = "states", values = "diff_4yr")+
  facet_wrap(~raceYear) +
  scale_fill_gradient2(low = "blue", mid="white",high = "red", 
  name = "Difference Scale")+labs(title='Difference in GOP 4-yr voteshare margin',
 subtitle='Red=More Republican, Blue=More Democrat, Gray=No Data')+
  theme(legend.position="right")+
  theme_void()

print(p_4yr)
```

As this map shows, there can be substantial variation in state-wide House 
popular vote share over years. Most of the largest variations (or darkest colors)
occur in the lower-populated states with fewer representatives, particularly when
at least one race is uncontested.

The two states that seemed to have lost their battleground status are Missouri and Colorado, with the former turning more red and the latter turning more blue. On the other hand, states slightly north like Michigan, Wisconsin,
Pennsylvania, and Minnesota appear to have stayed as battleground states in spite
of demographic and polarization changes. 

Next, I plotted the two year differences by state because our interest this year
is in predicting a midterm year. In this section, I only used maps from 1950 because
1948 was the first year in this dataset. 

```{r, message=FALSE}

# 4- year differences
p_2yr<-plot_usmap(data=h_avg_2%>% filter(raceYear>1948), regions = "states", values = "diff_2yr")+
  facet_wrap(~raceYear) +
  scale_fill_gradient2(low = "blue", mid="white",high = "red", 
  name = "Difference Scale")+labs(title='Difference in GOP 2-yr voteshare margin',
 subtitle='Red=More Republican, Blue=More Democrat, Gray=No data')+
  theme(legend.position="right")+
  theme_void()

print(p_2yr)
```

Similar to the earlier plot, the figure above shows that Democrats are losing
some of the Midwestern states while picking up some of the Southeast and Southwest.
Many of these states would still be considered battleground, but if patterns persist, 
it's likely that they would move to the 'lean/solid' category. 

I found it particularly interesting how a few recent midterm elections - 1994,
2010, 2014, and 2018- were associated with significant declines in the incumbent's
party's overall popular vote in house races by state. Similarly, I thought it was
noteworthy that the overall range in differences was much smaller for this plot. 
I believe this illustrates how an incumbent can often become popular after their
first term, and therefore capture even more of their vote in subsequent elections. 

*Results-Ohio*

Next, I looked at the specific district I'll be following - Ohio 01 - over time.

```{r, message=FALSE}

OH01<-h %>% filter(district_id=='OH01')%>%
  mutate(midterm=ifelse(raceYear%%4,0,1))
ohio_plot<-ggplot(data=OH01%>% filter(R_vote_margin>-.5),aes(x=raceYear,y=R_vote_margin, shape=WinnerParty,color=WinnerParty))+geom_point()+
  geom_textvline(label = "Steve Chabot enters", xintercept = 1994, vjust = -.4,linetype=2,color='red',size=3) +  facet_wrap(~RepStatus)+
  scale_color_manual(values=c("Blue","Red"))+
  ggtitle('Republican Vote Margin in OH-01 by Year and \nStatus of Republican Candidate (Challenger or Incumbent)')
print(ohio_plot)

```

Through this plot, I pay particular attention to voting patterns throughout the time 
since the incumbent, Steve Chabot, took office. As you can see, the district has 
been red the majority of the time, with incumbent Republicans doing especially well.

Steve Chabot first took office in 1994 and has been reelected every year but 2008. 
This could've been because of the Great Recession and overall discontent with the 
Republican party. While Chabot's popularity was rising throughout the mid 2010's,
it appears to have taken a hit with his winning margin only around .05 throughout the last 4 years. Another noteworthy element: I removed the data from 1992 where a Democrat (David Mann) ran uncontested and won because it was enlarging the y-axis too much. 
Mann ended up losing to Steve Chabot in 1994 by a sizable margin of 0.1. 


*Conclusion*

In my first blog post, I analyzed differences in state-wide house voting by 
two and four year intervals. While many battleground states remain in place like
Florida, Michigan, Wisconsin, some of the other states in the Midwest and Southwest
appear to be leaning more one party than earlier. 

As mentioned earlier, the popular vote does not carry a 1:1 relationship with 
seat share, so this blog is more examining trends than exact ramifications.

I look forward to updating this blog and diving deeper into the Ohio-01 district race. 
