---
title: 'Blog Post #2 - Economy'
author: Ellie Grueskin
date: '2022-09-15'
output:
  html_document: default
  pdf_document: default
slug: []
categories: []
summary: In this post, I look at the relationship between economic forces and US house voting patterns at the state and national level. As a fundamental part of the campaign environment, the a strong economy can help an incumbent president maintain power in other chambers while a gloomy economy can make the case harder for incumbents.
tags: []
---

# Intro

In this blog post (Extension #3), I  am examining how unemployment rates and labor force
participation rates affect voting patterns for House races in the United States at the state level.
I chose to model my outcome variable as the popular vote for House races by state, 
as opposed to seat share, because I thought this measure captured more of the nuance behind
each race. However, modeling seat share would be equally interesting and worthwhile to study.

# Data

For my local dataset, I join three datasets: Economic data by by state, popular voting by house (state level), and a national
dataset with presidential incumbent information. I restricted the economic data to just 
be in the 3rd quarter so it was the most recent for elections.
To compare local to national data, I built a national dataset of nationwide employment
and house voting. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(lubridate)
library(stargazer)
library("rmarkdown")
library("tinytex")
library("stargazer")
library("sandwich")
library('patchwork')
library('jtools')
library('huxtable')
library(usmap)
library(geomtextpath)



# Read state unemployment and house data 
unemp<-read.csv("data/unemployment_state_monthly.csv")
h_seat<-read.csv("data/house_popvote_seats.csv")%>% select(year,president_party)%>%
  mutate(Incumbent_Pres=lag(president_party))

# Read and reformat house popular vote
h_vote<-read.csv("data/house party vote share by district 1948-2020.csv")%>%
  select(raceYear, State,  RepVotes, DemVotes) %>%
# summarize party vote share by state
  group_by(State,raceYear) %>%
# mutate Rep vote margin by state %>%
  summarise(R_votemargin_st = (sum(RepVotes))/
           sum(RepVotes + DemVotes),
D_votemargin_st = (sum(DemVotes))/
sum(RepVotes + DemVotes)) %>%
rename(state = State)

# Clean Unemployment data
unemp_2 <- unemp %>% 
  filter(!(FIPS.Code%in% c(11,51000)|State.and.area=='Los Angeles County'))%>% 
  mutate(ElectionYear=ifelse(Year%%2,0,1))%>%
  filter(ElectionYear==1)%>%
  filter(Month==10|(Month==5 & Year==2022))%>%
  rename('state'='State.and.area')

#Join datasets together
dat<-left_join(unemp_2,h_vote,by=c('Year'='raceYear','state'))%>%
  filter(!R_votemargin_st%in% c(0,1))%>%
  mutate(state=factor(state))
dat<-left_join(dat,h_seat,by=c('Year'='year'))

# Clean data
dat_clean<-dat %>% group_by(state)%>%
  # Add lag variable and calculate Employment rate differences
  mutate(lag_R_vote=lag(R_votemargin_st),
         LFPR_diff=LaborForce_prct-lag(LaborForce_prct),
         Unemp_diff=(Unemployed_prct-lag(Unemployed_prct)),
         MidtermYear=ifelse(Year%%4,1,0)
         )

  # Add 2022 president party  
         # president_party=case_when(is.na(president_party)~'D',
         #             TRUE~president_party))

# Assign training and testing datasets
train<-dat_clean %>% filter(Year<2000)
test<-dat_clean %>% filter(Year>2000)
```

# Exploratory Plots
Using every state*election year from 1976-now, I look at trends 
for examining the state-wide Republican voteshare by differences in labor force
participation rate or unemployment rates between the past and current election.

```{r Exploratory plots,message=FALSE,warning=FALSE}
unemp_plot<-ggplot(data=dat_clean%>% filter(!is.na(Incumbent_Pres)),
       aes(x=Unemp_diff,y=R_votemargin_st,col=Incumbent_Pres))+
  geom_point(alpha=.4)+geom_smooth()+facet_wrap(~Incumbent_Pres)+ scale_color_manual(values=c("Blue","Red"))+
   theme(legend.position = "none")
unemp_midterm<-ggplot(data=dat_clean%>% filter(!is.na(Incumbent_Pres)&MidtermYear==1),
       aes(x=Unemp_diff,y=R_votemargin_st,col=Incumbent_Pres))+
  geom_point(alpha=.4)+geom_smooth()+facet_wrap(~Incumbent_Pres)+ scale_color_manual(values=c("Blue","Red"))
lfpr<-ggplot(data=dat_clean%>% filter(!is.na(Incumbent_Pres)),aes(x=LFPR_diff,y=R_votemargin_st,col=Incumbent_Pres))+geom_point(alpha=.4)+geom_smooth()+facet_wrap(~Incumbent_Pres)+
  scale_color_manual(values=c("Blue","Red"))+theme(legend.position='none')
lfpr_midterm<-ggplot(data=dat_clean%>% filter(!is.na(Incumbent_Pres)&MidtermYear==1),aes(x=LFPR_diff,y=R_votemargin_st,col=Incumbent_Pres))+geom_point(alpha=.4)+geom_smooth()+facet_wrap(~Incumbent_Pres)+
  scale_color_manual(values=c("Blue","Red"))
(unemp_plot+unemp_midterm)+plot_annotation(title='Republican Vote Share vs Difference in Unemployment Rate (2-year intervals)',subtitle='Left: All years, Right: Midterm Years, Both: Faceted and colored by Incumbent President Party')
(lfpr+lfpr_midterm)+plot_annotation(title='Republican Vote Share vs Difference in Labor Force Participation Rate (2-year intervals)',subtitle='Left: All years, Right: Midterm Years, Both: Faceted and colored by Incumbent President Party')
```
As these graphs show, there is not a obvious trend in any direction between these two
economic measures and Republican vote share by state.
For the times where the incumbent president was Republican, I expected there to be a positive relationship between labor force participation rate (LFPR) and Republican vote share along with a negative relationship between unemployment rate changes and Republican vote share. If the incumbent president was Democratic, I expected those relationships in inverse.
The lack of relationship could be due to the labor data being pooled at a state-wide level, while the races themselves represent smaller districts. Similarly, this could speak to the fact that these measures of the economy
are less tangible than indices like Real Disposable Income and CPI.

```{r national,include=FALSE}

unemp_nat<-read.csv("data/unemployment_national_quarterly_final.csv")
unemp_nat<-unemp_nat %>% 
  filter(quarter_cycle==3|(quarter_cycle==2 & year==2022))

h_seat<-read.csv("data/house_popvote_seats.csv")%>% 
  mutate(Incumbent_Pres=lag(president_party))

nat_vote<-h_seat%>%
  select(year,  R_votes, D_votes,Incumbent_Pres) %>%
# summarize party vote share by state
  group_by(year,Incumbent_Pres) %>%
# mutate Rep vote margin by state %>%
  summarise(R_votemargin = R_votes/(R_votes+D_votes))
joined_nat<-inner_join(nat_vote,unemp_nat,by='year')%>%
  mutate(lag=lag(R_votemargin),
         Unemp_diff=(UNRATE-lag(UNRATE)))
dat_nat<-inner_join(unemp_nat,nat_vote,by='year')
dat_nat<-dat_nat%>%
  mutate(lag_R_vote=lag(R_votemargin),
         Unemp_diff=(UNRATE-lag(UNRATE))/lag(UNRATE))

# Set train and test datasets
train_nat<-dat_nat %>% filter(year<2000)

```

# Modeling

To test these hypotheses further, I ran more rigorous models examining Republican
vote share by state against a variety of variables. I split my data into two datasets:
all elections before 2000 went into 'training,' or building the model, and and all
elections 2000 or later went to testing for checking my model. 

In the first regression, I 
use these independent variables: the state's Republican vote share from 2 years prior along with an interaction
for the party of the incumbent president with the difference in 
unemployment rates between the current and former election (time two years prior). 
In the second regression, I remove the state's vote share from two years prior, and in the third regression 
I use the same model as the first but restrict the data to only midterm years.

For models 4-6, I repeat the same steps except substitute the difference in 
Labor Force Participation Rate for the difference in unemployment rate. 
In models 6-7, I use national unemployment and voting data.

```{r results='asis'}

lm1<-lm(R_votemargin_st~lag_R_vote+Unemp_diff*factor(Incumbent_Pres),data=train)
lm2<-lm(R_votemargin_st~Unemp_diff*factor(Incumbent_Pres),data=train)
lm3<-lm(R_votemargin_st~lag_R_vote+Unemp_diff*factor(Incumbent_Pres),data=train%>% filter(MidtermYear==1))
lm4<-lm(R_votemargin_st~lag_R_vote+LFPR_diff*factor(Incumbent_Pres),data=train)
lm5<-lm(R_votemargin_st~LFPR_diff*factor(Incumbent_Pres),data=train)
lm6<-lm(R_votemargin_st~lag_R_vote+LFPR_diff*factor(Incumbent_Pres),data=train%>% filter(MidtermYear==1))
lm_nat<-lm(R_votemargin~lag_R_vote+Unemp_diff*factor(Incumbent_Pres),data=train_nat)
lm_nat_2<-lm(R_votemargin~Unemp_diff*factor(Incumbent_Pres),data=train_nat)

export_summs(lm1,lm2,lm3,lm4,lm5,lm6,lm_nat,lm_nat_2,
             model.names=c('R_Vote Share', 'R_Vote Share','R Vote Share (Midterm)',
                           'R_Vote Share', 'R_Vote Share','R Vote Share (Midterm)',
                           'R Vote Share (National)','R Vote Share (National)'))

```

From the first three columns, we see that the model with the highest R-squared,
or deviance explained, is the first column, which includes all years and information
about the state's vote in the election prior. 

The lag_R_vote should be interpreted as: every .01 point increase in the 
state's prior Republican vote is associated with a .0078 increase in the following
election's Republican vote. I refer to these by hundredths because the voting share is measured
in hundredths. This technique can be referred to
as serial correlation, where I am using the lagged variable of interest as an independent variable.

In all of the models, the coefficient for incumbent president is significant,
ranging from about -.03 to -.07. This suggests that a party's house vote is 
expected to decrease if it is the same party as the incumbent, even holding
constant state's voting patterns. It is interesting how the coefficients are larger
in absolute value in columns 3 and 6 because this may confirm prior studies showing
that midterm years are particularly bad for parties of the incumbent president.

Interestingly, the unemployment RATE difference term is not significant at the 5% level 
for incumbent Democrats (our base group), but has a very minor significant coefficient
in the context of Republican incumbent Presidents. 

Similarly, in my columns 4-6 where I repeat the same steps using LFPR as the
economic independent variable, the coefficients on the difference in LFPR between election years do not report significance at the 5% level and
the Republican vote share seems primarily dependent on the state's most 
recent voting patterns and the party of the Incumbent President.

The R-sq is significantly stronger with 
state-level data than nation-level data, however these models are challenging to compare
because there are vast differences in sample sizes.

# Model Validation

Next, I wanted to check how well my model had done, so I compared results between 
my predictions and the actual Republican vote shares in all years before 2022 
(since we don't have actual results for that yet). To save time, I only tested 
my first model, although testing the other would also prove fruitful.

```{r edit, include=FALSE}
test$predict<-predict(lm1,test)
test$predict2<-predict(lm2,test)
test$predict3<-predict(lm3,test)
test$predict4<-predict(lm4,test)
test$predict5<-predict(lm5,test)
test$predict6<-predict(lm6,test)

test_pre22<-test %>% filter(Year<2022)%>% mutate(
  error=predict-R_votemargin_st
)
```

```{r plots, message=FALSE}
ggplot(test_pre22,aes(x=R_votemargin_st,y=predict))+geom_point()+labs(title='Predictions vs Actual Values')
ggplot(test_pre22,aes(x=Year,y=error,color=factor(Incumbent_Pres)))+geom_point()+
  labs(title='Prediction Error By Year and Incumbent Party',color='Incumbent \n President Party')+
  scale_color_manual(values=c("Blue","Red"))
plot_usmap(data=test_pre22,regions='states',values='error')+facet_wrap(~Year)+
 scale_fill_gradient2(low = "red", mid="white",high = "blue", 
  name = "Difference Scale")+labs(title='Difference between Predictions and Acutal Values',
 subtitle='Blue=Results were more Democratic than Predicted, Red=Results were more Republican than Predicted')+
  theme(legend.position="right")

c(mean((lm1$model$R_votemargin_st-lm1$fitted.values)^2),
     mean((lm2$model$R_votemargin_st-lm2$fitted.values)^2),
     mean((lm3$model$R_votemargin_st-lm3$fitted.values)^2),
     mean((lm4$model$R_votemargin_st-lm4$fitted.values)^2),
     mean((lm5$model$R_votemargin_st-lm5$fitted.values)^2),
     mean((lm6$model$R_votemargin_st-lm6$fitted.values)^2))

```

These plots above show the distributions of errors across party, year, and state.
Fortunately, there does not appear to be a significant bias in my model's errors. 
However, to fully analyze the modeltake much more time and research to fully ana

Additionally, at the very bottom, I printed out the MSE for each local model. This shows
that including the lagged Republican vote share significantly reduces error, 
however it may be susceptible to overfitting in the long term.

# Ohio-01 Race

Finally, I looked at my congressional race- Ohio 01. Since this dataset is very small,
I was unable to find any meaningful patterns but it provided a little more background
into the economic landscape and voting patterns in the district.



```{r Ohio data, include=FALSE}
oh_vote<-read.csv("data/house party vote share by district 1948-2020.csv")%>%
  filter(State=='Ohio' & Area=='District 1')
oh_vote$RepStatus[which(oh_vote$raceYear == "2010")] <- 'Challenger'
# mutate Rep vote margin by state %>%
oh_vote<-oh_vote %>%
  group_by(raceYear,RepStatus)%>%
  summarise(R_votemargin = (sum(RepVotes))/
           sum(RepVotes + DemVotes),
D_votemargin = (sum(DemVotes))/
sum(RepVotes + DemVotes))%>%
  mutate(Incumbent_Votes=case_when(
    RepStatus=='Incumbent'~R_votemargin,
    RepStatus=='Challenger'~D_votemargin
  ))

unemp_ohio<-unemp %>% filter(State.and.area=='Ohio')%>%
  mutate(ElectionYear=ifelse(Year%%2,0,1))%>%
  filter(ElectionYear==1)%>%
  filter(Month==10|(Month==5 & Year==2022))%>%
  rename('state'='State.and.area')
dat_ohio<-left_join(unemp_ohio,oh_vote,by=c('Year'='raceYear'))
```

```{r ohio plots, message=FALSE}
ggplot(data=dat_ohio%>% filter(R_votemargin>0),aes(x=Year,y=Incumbent_Votes,
                    col=(Unemployed_prct-lag(Unemployed_prct))))+geom_point()+
  labs(color='Unemp  Diff',title='Incumbent Vote Share by Year and Unemployment difference between elections',x='Incumbent Vote Share') +
  geom_textvline(label = "Steve Chabot enters", xintercept = 1994, vjust = -.4,linetype=2,color='red',size=3)

ggplot(data=dat_ohio%>% filter(R_votemargin>0),aes(x=Year,y=Incumbent_Votes,col=(LaborForce_prct-lag(LaborForce_prct))))+geom_point()+scale_color_gradient(low='black',high='green')+labs(color='LFPR Diff',
          title='Incumbent Vote Share by Year and LFPR difference between elections',y='Incumbent Vote Share')+
   geom_textvline(label = "Steve Chabot enters", xintercept = 1994, vjust = -.4,linetype=2,color='red',size=3) 
```

# 2022 Predictions

To end my blog, I plotted my predictions for house vote share by state in 2022, using model #1.
Since I didn't yet have Q3 economic data, I used the latest month - May.
```{r predictions, message=FALSE}
test<- test %>% mutate(Incumbent_Pres=case_when(
                       Year==2022~'D',
                       TRUE ~ Incumbent_Pres))
test$predict<-predict(lm1,test)


p<-plot_usmap(data=test %>% filter(Year==2022),regions='states',values='predict',labels=TRUE)+
 scale_fill_gradient(low = "blue",high = "red")+
  labs(title='Predictions for 2022 Republican Vote Share by state',
       'Red= Strong R, Blue/Dark Purple= Strong D, Light Purple/Pink=Less clear')
p$layers[[2]]$aes_params$size <- 2
print(p)

```

# Conclusion

In this blog post, I explored the relationships between unemployment rate/ 
labor force participation rate and voting at the state level, and found very little 
significance of a clear correlation. Hence, this project would require further delving into
to see whether more nuanced models would detect a clear relationship or if 
the employment part of the economy is not what voters focus on.

My regressions did however show that voting is very serially correlated - 
districts have a strong correlation from year to year. Additionally, the incumbent
president plays a negative role, on average, particularly during midterm years.

Looking forward to 2022, this project is very relevant. Biden's economy has been
known for strong job growth but also overwhelming inflation. So, the economic 
measures we use to model outcomes are not necessarily substitutable and may
have grand implications for house popular vote/seat share.