---
title: Ground Game
author: R package build
date: '2022-10-16'
slug: []
categories: []
summary: Examining Ground Game
tags: []
---


# Intro

In this blog post, I look into the "Ground Game," or the non-ad components to political US House campaigns.
Primarily, I focus on Extension #2 whether expert predictions (blog post 4) and ad spending (blog post 5) 
affect turnout at the district level. I end with an update to my model and a further look
into my specific district, OH-01. 


```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE,warning=FALSE,message=FALSE)
library(tidyverse)
library(readr)
library(Metrics)
library(stargazer)

setwd("/Users/elliegrueskin/Documents/Fall 2022/Gov 1347/Gov1347/data")
exp<-read_csv("by_district/expert_rating.csv")%>% filter(year<2022)
inc<-read_csv("by_district/incumb_dist_1948-2020.csv")
cvap_district <- read_csv("by_district/cvap_district_2012-2020_clean.csv")
### District-wide data
cvap_district <- read_csv("by_district/cvap_district_2012-2020_clean.csv")%>%
  rename(st_cd_fips = geoid) 
inc_df<-read_csv("by_district/incumb_dist_1948-2020.csv")%>%
  filter(year == 2012 | year == 2014 | year == 2016 | year == 2018 | year == 2020)
expert<-read_csv("./by_district/expert_rating.csv")%>%
  filter(!is.na(district))%>%
  mutate(district_num=as.numeric(district),
         dist_num=case_when(
   district_num<10~paste0('0',district),
    TRUE~district
  ))
dist_pv_cvap <- inc_df %>%
  inner_join(cvap_district, by = c('st_cd_fips', 'year','state'))%>%
  mutate(district_num=as.numeric(district_num))
dat<-left_join(dist_pv_cvap,expert,by=c('year','state','district_num'))
dat <- dat %>%
  mutate(totalvotes = RepVotes + DemVotes,
         turnout = totalvotes/cvap,
         DemVotesMajorPct = DemVotesMajorPercent/100,
         RepVotesMajorPct = RepVotesMajorPercent/100)
dat_pred<-dat %>% filter(!is.na(avg_rating))

#make this example reproducible
set.seed(1)

#create ID column
dat_pred$id <- 1:nrow(dat_pred)

#use 70% of dataset as training set and 30% as test set 
train <- dat_pred %>% dplyr::sample_frac(0.70)
test  <- dplyr::anti_join(dat_pred, train, by = 'id')
```
# Ads

To begin, I examine whether there is a relationship between expert predictions
and turnout. I run three models, as shown below, with each model adding a variable.

In my first model, I simply run turnout, calculated as total voters minus the district's 
eligible citizens of that age group, on the average expert prediction rating. One immediate
limitation in this data is that there were only 390 observations for which I had expert
predictions between 2010-2022, so the training dataset becomes 270 observations.
The next model also includes a variable for the status of the incumbent, and the final 
model includes a variable for the presidential party.

```{r avg rating}

lm1<-lm(turnout~avg_rating,data=train)
lm2<-lm(turnout~avg_rating+RepStatus,data=train)
lm3<-lm(turnout~avg_rating+RepStatus+president_party,data=train)

accuracy<-function(model){
  test$pred<-predict(model,test,allow.new.levels=TRUE)
  rmse<-sqrt(mean(test$pred-test$turnout)^2)
  mae<-mae(test$pred,test$turnout)
  return(list('rmse'=rmse,'mae'=mae))
}


stargazer(lm1,lm2,lm3,type='text')
  
models<-list('lm1'=lm1,'lm2'=lm2,'lm3'=lm3)
tab<-sapply(models,accuracy)
tab
```

As the model output shows, my first model does not find there to be a significant
relationship between the turnout and average rating. However, after controlling for incumbency,
both models 2 and 3 determine that there is a negative relationship between the two variables.
This means that as Cook predicts that a district will turn more Republican, the turnout
is predicted to be about .01 points lower. Hence, the relationship is still about neglegible. 

Next, I created a small table with rmse and mae outputs as validation tests on my testing data.
Each of these measures is pretty similar, suggesting that even though our adjusted R^2 was much better after including incumbency data, the test data didn't show large differences.

To continue looking at turnout data, I examine the relationship between total $'s
spent on ads (for districts with data) and turnout in districts between 2010 and 2022.
This leads to 770 total observations.
As this first graph shows, there appears to be a positive, but noise, correlation
between ad spending and turnout, giving beginning evidence that ad spending 
is effective at mobilization. To control for midterm years, I next facet it by whether or not
the election was during a midterm year. I find that the R^2 is slightly better after
faceting, but the correlation is likely still hard to tease out without further data.
#ii. 
```{r ads}
ads<-read_csv("/Users/elliegrueskin/Documents/Fall 2022/Gov 1347/Gov1347/ads_joined_group.csv")
dat_ads<-inner_join(dist_pv_cvap,ads,by=c('year'='cycle','st_cd_fips'))
dat_ads <- dat_ads %>%
  mutate(totalvotes = RepVotes + DemVotes,
         turnout = totalvotes/cvap,
         DemVotesMajorPct = DemVotesMajorPercent/100,
         RepVotesMajorPct = RepVotesMajorPercent/100,
          MidtermYear=ifelse(year%%4,1,0) )
ggplot(data=dat_ads,aes(x=cost,y=turnout))+geom_point()

library(ggpmisc)
ggplot(data=dat_ads,aes(x=cost,y=turnout))+geom_point()+geom_smooth(method='lm')+stat_poly_eq()+facet_wrap(~MidtermYear,labeller=label_both)

```


Instead of going deeper into models for any of these relationships, I will return
to my existing model. Since these variables (turnout and ads) are hard to predict
ahead of time, I am going to stick to fundamental-based data in my model for now. 

This week, I will use last week's framework for the dataset, but instead modeling the outcome as the two party vote share. I will also include presidential approval polling.
```{r model}
inc<-read_csv("/Users/elliegrueskin/Documents/Fall 2022/Gov 1347/Gov1347/data/by_district/incumb_dist_1948-2020.csv")
inc_2<-inc %>%
  select(state,st_cd_fips, year, DemCandidate,RepCandidate,winner_party, RepStatus,DemStatus,RepVotesMajorPercent,DemVotesMajorPercent,president_party)%>%
  mutate(Uncontested=ifelse(DemVotesMajorPercent==0|RepVotesMajorPercent==0,1,0))

# Republican rows
inc_2_rep<-inc_2 %>% select(st_cd_fips, state,year, winner_party,president_party,Uncontested,RepCandidate, RepStatus,RepVotesMajorPercent)
# Dem rows
inc_2_dem<-inc_2 %>% select(st_cd_fips, state,year, winner_party,president_party,Uncontested,DemCandidate, DemStatus,DemVotesMajorPercent)
# Bind together
inc_2_total<-bind_rows(inc_2_rep,inc_2_dem)
inc_2_total<-inc_2_total %>%
  mutate(party=ifelse(is.na(RepVotesMajorPercent),'D','R'),
    CandidateName=ifelse(party=='R',RepCandidate,DemCandidate),
    Status=ifelse(party=='R',RepStatus,DemStatus),
    pv2p=ifelse(party=='R',RepVotesMajorPercent,DemVotesMajorPercent))%>%
  filter(pv2p!=0)%>%
  select(state,st_cd_fips, year,party,winner_party,president_party,Uncontested,
         party,CandidateName,Status,pv2p)


inc_2_total<-inc_2_total %>%
  mutate(party_wins=ifelse(party==winner_party,1,0))

# Unemployment data
unemp<-read_csv('/Users/elliegrueskin/Documents/Fall 2022/Gov 1347/Gov1347/data/by_state/unemployment_state_monthly.csv')

unemp_2 <- unemp %>% 
  filter(!('FIPS Code'%in% c(11,51000)|'State and area'=='Los Angeles County'))%>% 
  mutate(ElectionYear=ifelse(Year%%2,0,1))%>%
  filter(ElectionYear==1)%>%
  filter(Month==10)%>%
  rename('state'='State and area')

#Join datasets together
dat_total_minus_gen<-left_join(inc_2_total,unemp_2,by=c('year'='Year','state'))

# Generic polls
gen_poll<-read_csv('/Users/elliegrueskin/Documents/Fall 2022/Gov 1347/Gov1347/data/by_nation/polls_df.csv')
gen_poll_cl<-gen_poll %>% 
  filter(days_until_election<120)%>%
  group_by(year,party)%>%
  summarise(party_support=mean(support))

#Join datasets together
dat<-left_join(dat_total_minus_gen,gen_poll_cl,by=c('year','party'))%>%
  mutate(MidtermYear=ifelse(year%%4,1,0)) %>% 
mutate_if(is.numeric, ~replace(., is.na(.), 0))

dat_2<- dat %>%
  group_by(CandidateName)%>%
  arrange(year)%>%
 mutate(term = 1:n())%>%
  ungroup()


dat<-dat_2 %>%
  group_by(st_cd_fips,party)%>%
  arrange(year)%>%
  mutate(prev=lag(pv2p),Unc_prev=lag(Uncontested),
         approval_cent=party_support-44.6)%>%
  ungroup()
  
dat<- dat %>% mutate(same_party=ifelse(party==president_party,1,0))

lm1<-lm(pv2p~prev+Unc_prev+party_support+term+MidtermYear*same_party,data=dat %>% filter(Uncontested==0))
dat$pred<-predict(lm1,dat)
dat%>% filter(pred>50)%>%
  group_by(st_cd_fips,year)%>%
  summarise(n=n())%>% 
  filter(n==2)%>% view()

#ggplot(data=dat,aes(x=pv2p,y=predict(lm1,dat_2%>% filter(!is.na(pred[,"lwr"])))))
```





```{r,include=FALSE}
# library(readr)
# library(tidyverse)
# dist_pv_df <- read_csv("~Section data 4/incumb_dist_1948-2020 (3).csv")
# # read in cvap
# cvap_district <- read_csv("~/Dropbox/ElectionAnalytics_Midterms/Lab sessions/campaigns II_ ground game/cvap_district_2012-2020_clean.csv")
# # mutate geoid for merging
# cvap_district <- cvap_district %>%
#   rename(st_cd_fips = geoid)
# 
# # select relevant years from voting data
# table(dist_pv_df$year)
# # 2012 - from 2018
# # 2014, 2016, 2018, 2020 - from 2020
# dist_pv_df <- dist_pv_df %>%
#   filter(year == 2012 | year == 2014 | year == 2016 | year == 2018 | year == 2020)
# 
# table(dist_pv_df$st_cd_fips)
# 
# # merge
# dist_pv_cvap <- dist_pv_df %>%
#   inner_join(cvap_district, by = c('st_cd_fips', 'year'))
# 
# # mutate turnout
# dist_pv_cvap <- dist_pv_cvap %>%
#   mutate(totalvotes = RepVotes + DemVotes,
#          turnout = totalvotes/cvap)
# # save
# 
# # mutate votes percent for glm
# dist_pv_cvap <- dist_pv_cvap %>%
#   mutate(DemVotesMajorPct = DemVotesMajorPercent/100,
#          RepVotesMajorPct = RepVotesMajorPercent/100)
# 
# # drop uncontested seats
# dist_pv_cvap_closed <- dist_pv_cvap %>%
#   filter(!is.na(DemCandidate), !is.na(RepCandidate)) %>%
#   mutate(DemVotesMajorPct = DemVotesMajorPercent/100,
#          RepVotesMajorPct = RepVotesMajorPercent/100)
# 
# # basic lm
# fit1 <- lm(DemVotesMajorPercent ~ turnout,
#               data = dist_pv_cvap_closed)
# summary(fit1)
# 
# fit1_glm <- glm(DemVotesMajorPct ~ turnout,
#               data = dist_pv_cvap_closed, family = binomial(link="logit"))
# summary(fit1_glm)
# ```
# 
# 
# ```{r,include=FALSE}
# lm1<-lm(turnout~cost,data=dat_ads)
# 
# 
# dat_22<-read_csv("./by_district/house_cands.csv")
# dat_22<-dat_22 %>% filter(cand_party%in%c('Democrat','Republican'))
# 
# ggplot(dat_pred,aes(x=cook,y=DemVotesMajorPct))+geom_point()+geom_smooth(method='lm')
# lm1<-lm(turnout~I((avg_rating-4)^2),data=dat_pred)
# 
# setwd("/Users/elliegrueskin/Documents/Fall 2022/Gov 1347/Gov1347/data")
# exp<-read_csv("by_district/expert_rating.csv")%>% filter(year<2022)
# inc<-read_csv("by_district/incumb_dist_1948-2020.csv")
# cvap_district <- read_csv("by_district/cvap_district_2012-2020_clean.csv")
# cvap_district <- cvap_district %>%
#   rename(st_cd_fips = geoid) 
# dat_exp<-merge(exp,cvap_district,by=)
# 
# polls_df<-read_csv("Section data 4/house_polls_long.csv")
# polls_df <- polls_df %>%
#   filter(year == '2018' | year == '2020')
# table(polls_df$year)
# # join
# cvap_district <- cvap_district %>%
#   # filter to relevant years 
#   filter(year == '2018' | year == '2020')
# 
# polls_cvap_df <- merge(polls_df, cvap_district, by = c('st_cd_fips', 'year'))
# 
# dist_pv_df <- read_csv("Section data 4/incumb_dist_1948-2020 (3).csv")
# # select relevant years from voting data
# dist_pv_df <- dist_pv_df %>%
#   filter(year == 2012 | year == 2014 | year == 2016 | year == 2018 | year == 2020)
# 
# # merge
# dist_pv_cvap <- dist_pv_df %>%
#   inner_join(cvap_district, by = c('st_cd_fips', 'year'))
# polls_cvap_vp_df <- merge(polls_cvap_df, dist_pv_df, by = c('st_cd_fips', 'year'))
# 
# # mutate turnout
# polls_cvap_vp_df <- polls_cvap_vp_df %>%
#   mutate(totalvotes = RepVotes + DemVotes,
#          turnout = totalvotes/cvap,
#          DemVotesMajorPct = DemVotesMajorPercent/100,
#          RepVotesMajorPct = RepVotesMajorPercent/100)
# 
# # drop uncontested seats
# dist_pv_cvap_closed <- polls_cvap_vp_df %>%
#   filter(!is.na(DemCandidate), !is.na(RepCandidate)) %>%
#   mutate(DemVotesMajorPct = DemVotesMajorPercent/100,
#          RepVotesMajorPct = RepVotesMajorPercent/100)
# 
# # basic lm
# fit1 <- lm(DemVotesMajorPercent ~ turnout,
#               data = dist_pv_cvap_closed)
# summary(fit1)
# 
# ggplot(dist_pv_cvap_closed, aes(x = turnout, y = DemVotesMajorPercent)) + 
#   geom_point() +
#   stat_smooth(method = "lm")+
#   facet_wrap(~president_party*DemStatus,labeller=label_both)
# 
# inc<-read_csv("/Users/elliegrueskin/Documents/Fall 2022/Gov 1347/Gov1347/data/by_district/inc.csv")
# 
# 
# # join
# df <- left_join(inc, cvap_district, by = c('st_cd_fips', 'Year'='year','state'))
# 
# # merge with district-level voteshares
# 
# dat <- dat %>%
#   group_by(st_cd_fips,year)
#   mutate(totalvotes = RepVotes + DemVotes,
#          turnout = totalvotes/cvap)
# 
# ggplot(data=df,aes(x=cvap,y=pv2p,col=party))+geom_smooth()
# +facet_wrap(~Status,labeller=label_both)
```
