---
title: "Air War"
author: "R package build"
date: "2022-10-07"
slug: []
categories: []
summary: In this post, I look at the effects of campaign ads on predicting US house
  elections.
tags: []
---

# Intro
.. Over xx % of the average competitive campaign's budget goes to advertising...


# Data

My primary focus this week was reformatting my data to be in a longer format, such
that each row represented a candidate for each cycle's race.  I began by reformatting
the incumbency data and then joined that to data from state economic variables, 
generic ballots, and advertising information from the Wesleyan Media Project.

The advertising data refers to all the tv ads during a candidate's cycle. 


```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE,warning=FALSE,message=FALSE)
library(tidyverse)
library(lubridate)
library(stargazer)
library("rmarkdown")
library("tinytex")
library("stargazer")
library("sandwich")
library('patchwork')
library('jtools')
library('huxtable')
library(usmap)
library(geomtextpath)
library(caret)
library(pROC)

# Read in ad data and merge
# ads<-read_csv("/Users/elliegrueskin/Documents/Fall 2022/Gov 1347/Gov1347/data/air_war/ads_2006_2018 (1).csv")
# ads_issues<-read_csv("/Users/elliegrueskin/Documents/Fall 2022/Gov 1347/Gov1347/data/air_war/ads_issues_2012-2018 (1).csv")
# ads_joined<-left_join(ads,ads_issues,by=c('creative','market','dma','airdate','airtime',
#       "state","race","district","party","ad_tone","ad_purpose",  
#       "station","cycle" ,"st_fips","cd_fips","st_cd_fips"))
# ads_joined %>% filter(is.na(est_cost))%>% nrow()
# # Ads grouped by cycle, race, and party
# ads_joined_group<-ads_joined %>%
#   filter(!is.na(party))%>%
#   group_by(cycle,st_cd_fips,party)%>%
#   summarise(total_ads=n(),
#             stations=sum(n_stations),
#             attack_ad_pct=sum(ad_tone=='attack')/total_ads)%>%
#   mutate(party=substr(party,1,1))
ads_joined_group<-read_csv("/Users/elliegrueskin/Documents/Fall 2022/Gov 1347/Gov1347/data/by_district/ads_joined_group.csv")

# Read in incumbency, voting data and clean to wide format with data related to ads data

inc<-read_csv("/Users/elliegrueskin/Documents/Fall 2022/Gov 1347/Gov1347/data/by_district/incumb_dist_1948-2020.csv")
inc_2<-inc %>% filter(year>2004&year<2020) %>% 
  select(state,st_cd_fips, year, DemCandidate,RepCandidate,winner_party, RepStatus,DemStatus,RepVotesMajorPercent,DemVotesMajorPercent,president_party)%>%
  mutate(Uncontested=ifelse(DemVotesMajorPercent==0|RepVotesMajorPercent==0,1,0))

# Republican rows
inc_2_rep<-inc_2 %>% select(st_cd_fips, state,year, winner_party,president_party,Uncontested,RepCandidate, RepStatus,RepVotesMajorPercent)
# Dem rows
inc_2_dem<-inc_2 %>% select(st_cd_fips, state,year, winner_party,president_party,Uncontested,DemCandidate, DemStatus,DemVotesMajorPercent)
# Bind together
inc_2_total<-bind_rows(inc_2_rep,inc_2_dem)
inc_2_total<-inc_2_total %>%
  mutate(party=ifelse(is.na(RepVotesMajorPercent),'D','R'),
    CandidateName=ifelse(party=='R',RepCandidate,DemCandidate),
    Status=ifelse(party=='R',RepStatus,DemStatus),
    pv2p=ifelse(party=='R',RepVotesMajorPercent,DemVotesMajorPercent))%>%
  filter(pv2p!=0)%>%
  select(state,st_cd_fips, year,party,winner_party,president_party,Uncontested,
         party,CandidateName,Status,pv2p)


# Join to ads data
dat_total<-full_join(ads_joined_group,inc_2_total, by=c('cycle'='year','st_cd_fips','party'))%>%
  filter(!is.na(winner_party))
dat_total<-dat_total %>%
  mutate(party_wins=ifelse(party==winner_party,1,0))

# Unemployment data
unemp<-read_csv('/Users/elliegrueskin/Documents/Fall 2022/Gov 1347/Gov1347/data/by_state/unemployment_state_monthly.csv')

unemp_2 <- unemp %>% 
  filter(Year>2004&Year<2020) %>% 
  filter(!('FIPS Code'%in% c(11,51000)|'State and area'=='Los Angeles County'))%>% 
  mutate(ElectionYear=ifelse(Year%%2,0,1))%>%
  filter(ElectionYear==1)%>%
  filter(Month==10)%>%
  rename('state'='State and area')

#Join datasets together
dat_total_minus_gen<-left_join(dat_total,unemp_2,by=c('cycle'='Year','state'))

# Generic polls
gen_poll<-read_csv('/Users/elliegrueskin/Documents/Fall 2022/Gov 1347/Gov1347/data/by_nation/polls_df.csv')
gen_poll_cl<-gen_poll %>% 
  filter(days_until_election<120)%>%
  group_by(year,party)%>%
  summarise(per=mean(support))

#Join datasets together
dat<-left_join(dat_total_minus_gen,gen_poll_cl,by=c('cycle'='year','party'))%>%
  mutate(MidtermYear=ifelse(cycle%%4,1,0)) %>% 
mutate_if(is.numeric, ~replace(., is.na(.), 0))%>%
  mutate(ads_run=ifelse(total_ads>0,1,0))


dat<-dat %>%
  group_by(CandidateName)%>%
  arrange(cycle)%>%
  mutate(term=ifelse(Status=='Incumbent',cumsum(n()),0))%>%
  ungroup()

dat<-dat %>%
  group_by(st_cd_fips,party)%>%
  arrange(cycle)%>%
  mutate(prev=lag(pv2p),Unc_prev=lag(Uncontested),
         approval_cent=per-44.6)%>%
  ungroup()
  
rm(ads)
rm(ads_issues)

```

```{r EDA}
dat_total %>% filter(cycle==2018&total_ads>0)%>%
  ggplot(aes(x=total_ads,y=pv2p,size=stations))+geom_point()+
  facet_wrap(~party) + labs(title='Popular vote vs. total ads run, 2018')

dat_total<- dat_total %>% mutate(ads_comb=total_ads*stations)

dat_total %>% filter(pv2p>45&pv2p<55&cycle==2018&total_ads>0)%>%
  ggplot(aes(x=total_ads,y=pv2p,size=stations))+geom_point()+
  facet_wrap(~party)+labs(title='Popular vote vs. total ads run, 2018')

dat_total %>% filter(ads_comb>quantile(dat_total$ads_comb,.7,na.rm=T)&(pv2p>60|pv2p<40))%>% 
  ggplot(aes(x=total_ads,y=pv2p,col=factor(party)))+geom_point()
dat %>% filter(Uncontested==0&Unc_prev==0) %>%ggplot(aes(x=prev,y=party_wins,col=winner_party))+
  geom_smooth()

```


# Model

```{r model}

# previous years 
dat_test<-dat %>% filter(cycle=='2018'&Uncontested==0)
dat_train<-dat %>% filter(cycle<'2018'&Uncontested==0)
g0<-glm(party_wins~prev:factor(Unc_prev)+
          party:president_party+factor(MidtermYear)+Status+term,
        family=binomial(link=logit),data=dat_train)
g1<-glm(party_wins~prev:factor(Unc_prev)+
          Unemployed_prct+
          party:president_party+factor(MidtermYear)+Status+term+(per)+
          ads_run+total_ads,family=binomial(link=logit),data=dat_train)

g2<-glm(party_wins~prev:factor(Unc_prev)+
          Unemployed_prct+
          party:president_party+factor(MidtermYear)+Status+term+stations+
          ads_run+total_ads,family=binomial(link=logit),data=dat_train)
g2<-glm(party_wins~prev*factor(Unc_prev)+party:president_party:MidtermYear+term+Status+factor(Uncontested),family=binomial(link=logit),data=dat%>% filter(cycle<2018))

g1<-glm(party_wins~party:president_party:MidtermYear+Status+term,family=binomial(link=logit),data=dat%>% filter(cycle<2018))
g2<-glm(party_wins~party:president_party:MidtermYear+Status+term+ads_run+total_ads+Unemployed_prct,family=binomial(link=logit),data=dat%>% filter(cycle<2018))
ggplot(data=dat_train,aes(x=predict(g0,dat_train,type='response'),y=pv2p))+geom_smooth()


ggplot(data=dat_train,aes(x=predict(g2,dat_train,type='response'),y=party_wins))+geom_smooth()
# AUC/Sensitivity/Specificity Function

classification_table<-function(model){
  dat_test$pred<-predict(model,dat_test,type='response')
  classificationtable<-table(dat_test$party_wins,dat_test$pred > .5)
  sensitivity<-(classificationtable[2,2]/(classificationtable[2,2]+classificationtable[2,1]))*100
  specificity<-(classificationtable[1,1]/(classificationtable[1,1]+classificationtable[1,2]))*100
  auc<-roc(dat_test$party_wins,dat_test$pred)
  return(list('auc'=auc$auc,'sensitivity'=sensitivity,'specificity'=specificity))
}

models<-list('log1'=g1,'log2'=g2)
tab<-sapply(models,classification_table)




```

# Results 


# Extra
```{r}
# 
# # Pres polls
# pres_poll<-read_csv('/Users/elliegrueskin/Documents/Fall 2022/Gov 1347/Gov1347/data/by_nation/approval_polllist.csv')
# pres_poll_cl<-pres_poll%>% mutate(election_date=as.Date(election_date,format = "%m/%d/%y"),
#                                   last_enddate=as.Date(last_enddate,format = "%m/%d/%y"))%>%
#   mutate(days_until_election=difftime(election_date,last_enddate,units="days"))%>%
#   filter(days_until_election<120)%>%
#   filter(cycle>2000&!(state%in% c('National','District of Columbia')))%>%
#   mutate(state=case_when(
#     state%in% c('NE-1','NE-2','NE-3')~'Nebraska',
#     state%in% c('ME-1','ME-2')~'Maine',
#     TRUE~state
#   ))%>%
#   group_by(candidate_id,cycle,state,candidate_name)%>%
#   summarise(approval=mean(pct_estimate))
#     
```


# Conclusion

Moving forward, I don't plan to use ads data because there isn't a comprehensive database
for 2022 candidates. However, I think this exploration provided really interesting insight, 
beginning to suggest that advertising has a positive significant impact on a candidate's
odds of getting elected. 

# Citations

Alan S Gerber, James G Gimpel, Donald P Green, and Daron R Shaw. How Large and Long- lasting are the Persuasive Effects of Televised Campaign Ads? Results from a Randomized Field Experiment. American Political Science Review, 105(01):135–150, 2011.
